---
title: "Class 01: An introductory data wrangling and description example"
author: "Ethan/Gutknecht"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preliminaries

Before getting started we need to give credit where credit is due. A special thanks to Mine Cetinkaya-Rundel from Duke University for the idea and outlines of this first assignments.

## Opening question

Today we will look at the voting history of countries in the United Nations General Assembly. We will be using data from the `unvotes` package. Additionally, we will make use of the `tidyverse` and `lubridate` packages.

```{r load-packages, warning=FALSE, message=FALSE}
library(unvotes)
library(tidyverse)
library(lubridate)
```

## Data

We have three datasets we can work with: `un_roll_calls`, `un_roll_call_issues`, and `un_votes`. Each of these datasets contains a variable called `rcid`, the roll call id, which can be used to join them with each other.

- The `un_votes` dataset provides information on the voting history of the United Nations General Assembly. It contains one row for each country-vote pair.

```{r view-un_votes}
un_votes
```

- The `un_roll_calls` dataset contains information on each roll call vote of the United Nations General Assembly.

```{r view-un_roll_calls}
head(un_roll_calls)
```

- The `un_roll_call_issues` dataset contains (topic) classifications of roll call votes of the United Nations General Assembly. Many votes had no topic, and some have more than one.

```{r view-un_roll_call_issues}
tail(un_roll_call_issues)
```

## Analysis

### Larger comparison example

First, let's take a look at how often each country voted "yes" on a resolution in each year. We'll visualize the results, so let's pick a few countries of interest first,

```{r country-list}
country_list <- c("United States", "China", "Mexico", "Canada")
```

and focus our analysis on them. In this first *chunk* of code we bring together the `un_votes` dataset with the `un_roll_calls` dataset linking by the `rcid` variable (this is the `inner_join` call). The `filter()` statement limits our data to the countries of interest. The `group_by()` and `summarize()` functions calculate the number of `votes` and the percentage of times a country voted "yes" for each country per year. Lastly we take a sneak peak at our summarized data with the `glimpse()` function.

```{r clean-yearly-yes, fig.width=10, warning=FALSE}
un_vote.trim <- un_votes %>%
  filter(country %in% country_list) %>%
  inner_join(un_roll_calls, by = "rcid") %>%
  group_by(Year = year(date), country) %>%
  summarize(votes = n(),
            percent_yes = mean(vote == "yes"))
glimpse(un_vote.trim)
```

Now that we have the data as we way, we can plot it. In the code below note that we are letting the Country determine the color of the line.

```{r plot-yearly-yes, fig.width=10, warning=FALSE, fig.cap="Figure 1: UN voting records over time for 4 countries. The four lines are close together until the late 1960s, but after that they gradually split with Mexico at about 90%, China at about 80%, Canada at about 50%, and the US at about 20% over the last 50 years."}
ggplot(un_vote.trim, aes(x = Year, y = percent_yes, color = country)) +
  geom_line() +
  ylab("% of votes that are 'Yes'")
```

#### Discussion

What do you notice about the difference in the countries?


### Comparison of two countries

Next, let's see how *United States* and *China*'s voting record changed over the years on each of the issues.

```{r plot-yearly-yes-issue, fig.width=10, warning=FALSE, fig.cap="Figure 2: Plots of UN voting records for the United States and China on the following 6 issues over time: arms control, colonialism, economic development, human rights, nuclear weapons, and Palestinian conflict. The lines on the plot show that China consistently votes yes at a much higher rate than the United States."}
un.compare <- un_votes %>%
  filter(country %in% c("United States", "China")) %>%
  inner_join(un_roll_calls, by = "rcid") %>%
  inner_join(un_roll_call_issues, by = "rcid") %>%
  group_by(country, Year = year(date), issue) %>%
  summarize(votes = n(),
            percent_yes = mean(vote == "yes")) %>%
  filter(votes > 5)    # Only use records where there are more than 5 votes

ggplot(un.compare, aes(x = Year, y = percent_yes, color = country)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ issue) +
  ylab("% of votes that are 'Yes'")
```


## Now it's your turn!!!

### Part 0

Change the line at the top of this file labeled "author:" to match your name and those of your teammates.

### Part 1

Pick two other countries (the full list is below) besides the *United States of America* and *China*. Write in which two countries you chose and why you picked them. 

Australia and United Kingdom. They are bigger countries and most likely will have a bigger influence on world issues.

### Part 2

Edit the code in the "Comparison of two countries" part of this document to do the following:

* Compare the two countries you chose
* Filter the data such that it is only plotted if:
    - there was more than 10 votes
    - The `Year` is after the year 2000


```{r plot-yearly-yes-issue-part2, fig.width=10, warning=FALSE}
un.compare <- un_votes %>%
  filter(country %in% c("Australia", "United Kingdom")) %>%
  inner_join(un_roll_calls, by = "rcid") %>%
  inner_join(un_roll_call_issues, by = "rcid") %>%
  group_by(country, Year = year(date), issue) %>%
  summarize(votes = n(),
            percent_yes = mean(vote == "yes")) %>%
  filter(votes > 10, Year > 2000) # Only use records where there are more than 10 votes. Only use records where date is past 2000

ggplot(un.compare, aes(x = Year, y = percent_yes, color = country)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ issue) +
  ylab("% of votes that are 'Yes'")
```

### Part 3

Compare contrast the two countries you chose. Does anything seem surprising in their voting tendencies? Discuss!

All of the votes are very similar and have a direct correlation besides the Palestinian conflict. We can see that Australia's votes sways from the United Kingdoms.

### Part 4

*Knit* the document once more and submit the *html* file through the canvas site.


## References

1. David Robinson (2017). unvotes: United Nations General Assembly Voting Data. R package version 0.2.0. https://CRAN.R-project.org/package=unvotes.
2. Erik Voeten "Data and Analyses of Voting in the UN General Assembly" Routledge Handbook of International Organization, edited by Bob Reinalda (published May 27, 2013).
3. Much of the analysis has been modeled on the examples presented in the [unvotes package vignette](https://cran.r-project.org/web/packages/unvotes/vignettes/unvotes.html).

## Appendix

Below is a list of countries in the dataset:

```{r}
un_votes %>% 
  arrange(country) %>% 
  select(country) %>%
  distinct() %>%
  as.data.frame()
```




